# Introduction to graphics in R

```{r library_declaration, eval=T, message=FALSE, warning=T, include=FALSE, echo=F}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)
library(gapminder)

```

## Setup

The package suite `tidyverse` which includes the package `ggplot2` needs to be included.
For installing `ggplot2`, type `install.packages("ggplot2")` or equivalently for `tidyverse`,
type `install.packages("tidyverse")`

## Introduction to ggplot

The `gg` of `ggplot` stands for (layered) "grammar of graphics" [@Wilkinson:2005], [@Wickham:2010]. 
This idea will be further explored by the means of data from the package `gapminder()`. To install, type `install.packages("gapminder")` in the RStudio console.

```{r data_load, eval=T, message=F, warning=T, include=T, echo=T}

data_gapminder <- gapminder::gapminder 

```

### Notes

1. Why `<-` as opposed to `=` ?

2. Why `gapminder::gapminder` ?

3. What is a dataframe?

A data frame is a rectangular collection of variables (in the columns) and observations (in the rows). It's different from a 'mere' matrix since the columns have variable names usually and can contain different "types", say numeric, categorical, character and logical all together. 

## Questions

### Are "Western" countries richer than "Eastern" countries?

The current GDP/capita of Europe (in 2007):

```{r plot_eur, eval=T, echo=T, message=F, warning=T, include=T}

data_eur_2007 <- data_gapminder %>% #what is this funny sign?
  dplyr::filter(year == 2007) %>% #isolates variables for year 2007
  dplyr::filter(continent == "Europe") 

(plot_eur <- ggplot(data = data_eur_2007) +
  geom_point(mapping = aes(x = gdpPercap, y = country))
  ) #why parentheses around the command?

```

There is high variation---from Albania to Norway. 

What about Asian countries in 2007?

```{r plot_asia, eval=T, echo=T, message=F, warning=T, include=T}

data_asia_2007 <- data_gapminder %>% 
  dplyr::filter(year == 2007) %>% #isolates variables for year 2007
  dplyr::filter(continent == "Asia") #collect only Asian countries

(plot_asia <- ggplot(data = data_asia_2007) +
  geom_point(mapping = aes(x = gdpPercap, y = country)) +
  labs(x = "GDP/capita (USD)", 
       y = "Country",
       title = "GDP per capita in Asia",
       subtitle = "Put subtitle here"))

```

Again, large variation among Asian countries but what about the bounds? What can we say about the question?

### Graphics

We start with the function `ggplot()`. It creates a coordinate system that we will add layers to. The first argument is the dataset to use in the graph. 

`ggplot(data = data_eur_2007)` creates an empty graph. The function `geom_point()` adds a layer of points to our plot. Each geom function in ggplot2 takes a mapping argument. This defines how variables in our dataset are mapped to aesthetics such as axes, colors, shapes etc. The $x$ and $y$ arguments of `aes()` specify which variables to map to the $x$ and $y$ axes. Variables can also be mapped to aesthetics such as colors, shapes, sizes etc. 


### A More Granular Look: China, India, France, Germany

What about life expectancy in these countries with time?

```{r CIFG, message=F, warning=T, eval=T, echo=T, include=T}

data_CIFG <- data_gapminder %>%
  dplyr::filter(country %in% c("China", 
                               "India", 
                               "France", 
                               "Germany"
                               )
                )

(plot_CIFG_life_exp <- ggplot(data = data_CIFG) +
  geom_line(mapping = aes(x = year, 
                          y = lifeExp, 
                          color = country)
            )
  )

```

### Notes

1. Plots can be stored as variables too.
2. Other aesthetic attributes: shape, size, alpha (transparency) etc.
3. Note where to put the `+` sign.
4. `geom_line()` as opposed to points. Other "geoms" are `geom_smooth`, `geom_boxplot`, `geom_bar` etc.


### Faceting

```{r CIFG_Facet, warning=T, message=F, include=T, eval=T, echo=T}

(plot_CIFG_life_cont <- ggplot(data = data_CIFG) +
  geom_line(mapping = aes(x = year, 
                          y = lifeExp, 
                          color = country)
            ) +
  facet_wrap(~ continent)
 )

```

### Notes

1. To facet on one variable ('continent' here), use `facet_wrap()`. 
2. To facet on two variables, use `facet_grid()`

### The Notion of Geoms in `ggplot()`

Sometimes one needs bar charts, sometimes histograms; at other times scatterplots or lines etc. All these are different "geoms" in `ggplot()`. Scatterplots can be made with `geom_point()`, lineplots can be implemented with `geom_line()`; boxplots require `geom_boxplot()`, barcharts need `geom_bar()` and so on. Multiple geoms could be part of the same graph. The library ggplot2 provides over 30 geoms.

Each geom will need "aesthetic" parameters: for example, which datasets form the $x$ axis? Which ones form the $y$ axis? What colors to use for different variables? 

Somewhat unsurprisingly, not every aesthetic works with every geom. We could set the shape parameter of a point, but cannot do so for a line. 

This is how we could include multiple geoms in the same plot:

```{r CIFG_Mult_Geom, warning=T, message=F, include=T, echo=T, eval=T}
(plot_CIFG_mult_geom <- ggplot(data = data_CIFG) +
   geom_point(mapping = aes(x = year, y = lifeExp, color = country)) +
   geom_line(mapping = aes(x = year, y = lifeExp, linetype = country))
  )

```

Note however, that we need to rewrite the code for aesthetic parameters: `x = year, y = lifeExp`. We could re-express the same idea in fewer lines---by passing some common aesthetic parameters as global options---by including them in the main `ggplot` argument:

```{r CIFG_Mult_Geom_2, warning=F, include=T, eval=T, echo=T, message=F}

plot_CIFG_mult_geom_2 <- ggplot(data = data_CIFG, 
                                mapping = aes(x = year, 
                                              y = lifeExp)) +
  geom_point(mapping =  aes(color = country)) +
  geom_line(mapping = aes(linetype = country))
 

```

### Plotting linear trends: `geom_smooth()`

Fitting a line to a set of observations is linear smoothing. Fitting a polynomial of degree 2 is quadratic smoothing and so on. When we wish to plot observations and the best linear fit computed by linear regression, we need to use `geom_smooth()` with the option `method = "lm"` where `lm()` stands for "linear model".

### Question: What was population growth in India?

```{r CIFG_pop_Ind, include=T, echo=T, eval=T, message=F, warning=T}

(plot_CIFG_pop_Ind <- ggplot(data = filter(data_CIFG, #why no +?
                                           country == "India"), 
                         mapping = aes(x = year, y = pop)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) #se = "standard errors"
 )

```

### Question: What about the same in China?

```{r CIFG_pop_China, include=T, echo=T, eval=T, message=F, warning=T}

(plot_CIFG_pop_China <- ggplot(data = filter(data_CIFG, 
                                             country == "China"), 
                         mapping = aes(x = year, y = pop)) +
  geom_point() +
  geom_smooth(method = "lm") #se = T by default
 )

```

The default smoothing method however is "loess" (locally weighted scatterplot smoothing)

```{r CIFG_pop_Asia_loess, include=T, echo=T, eval=T, message=F, warning=T}

(plot_CIFG_pop_Ind <- ggplot(data = filter(data_CIFG, 
                                           continent == "Asia"), 
                         mapping = aes(x = year, y = pop)) +
  geom_point(mapping = aes(color = country)) +
  geom_smooth() 
 )

```

### Bar Charts

What about GDP per capita of CIFG countries in 2007?

```{r CIFG_pop_2007, include=T, echo=T, eval=T, message=F, warning=T}

(plot_CIFG_pop_2007 <- ggplot(data = filter(data_CIFG,
                                            year == 2007),
                              mapping = aes(x = country, 
                                            y = gdpPercap)) + 
   geom_bar(mapping = aes(fill = country), #what does this mean?
            stat = "identity") #what does this mean?
 )

```

While some types of plots (scatterplots) do not require any transformation---each point is plotted at $x$ and $y$ coordinates same as the original value, others (such as boxplots, histograms etc.) require transformations. For example, for boxplots, the median and the interquartile range (IQR) need computing.

### Boxplots

```{r plot_CIFG_box, include=T, echo=T, eval=T, message=F, warning=T}

(plot_CIFG_box <- ggplot(data = data_CIFG,
                         mapping = aes(x = country, 
                                       y = lifeExp,
                                       color = country)) +
   geom_boxplot()
   )

```

In some cases there are many variables for which we need boxplots; and including all of them on the $x$ axis may cause their names to squish together, leading to poor visibility. In such cases, one can use the command `coord_flip()` which, as the name suggests, flips the coordinates---from $x$ to $y$ and vice versa.

```{r plot_eur_box_flip, include=T, echo=T, eval=T, message=F, warning=T}

(plot_eur_box_flip <- ggplot(data = filter(data_gapminder,
                                           continent == "Europe"),
                         mapping = aes(x = country, 
                                       y = lifeExp,
                                       color = country)) +
   geom_boxplot() +
   coord_flip()
   )

```

### Themes

Themes govern the overall "look" of the plot. The default theme is `theme_grey()`. However, there are dozens of other themes which one may prefer. A small set of examples is presented here.

```{r CIFG_theme, warning=F, include=T, eval=T, echo=T, message=F}

(plot_CIFG_theme <- ggplot(data = data_CIFG, 
                                mapping = aes(x = year, 
                                              y = lifeExp)) +
  geom_point(mapping =  aes(color = country)) +
  geom_line(mapping = aes(linetype = country)) +
   labs(x = "Years", 
        y = "Life Expectancy",
        title = "Illustration of Different Themes") +
  theme_classic() #note the classic theme
 )
 
(plot_CIFG_theme <- ggplot(data = data_CIFG, 
                                mapping = aes(x = year, 
                                              y = lifeExp)) +
  geom_point(mapping =  aes(color = country)) +
  geom_line(mapping = aes(linetype = country)) +
   labs(x = "Years", 
        y = "Life Expectancy",
        title = "Illustration of Different Themes") +
  theme_bw() #note the bw theme (black and white)
 )

(plot_CIFG_theme <- ggplot(data = data_CIFG, 
                                mapping = aes(x = year, 
                                              y = lifeExp)) +
  geom_point(mapping =  aes(color = country)) +
  geom_line(mapping = aes(linetype = country)) +
   labs(x = "Years", 
        y = "Life Expectancy",
        caption = "Insert caption here",
        title = "Illustration of Different Themes") +
  theme_minimal() #note the minimal theme
 )

```



## FAQ: How to Plot Multiple Variables Together?

### Wide Versus Long Data

When each column is a separate variable, the dataset is said to be in the "wide" format. In the "long" format, variables are "gathered" together in one column with another containing values.



```{r wide_vs_long, warning=F, include=T, eval=T, echo=T, message=F}

temp_w <- cbind(1991:2010, 
                rnorm(20, 0, 1), 
                rnorm(20, 2, 5),
                rnorm(20, 1, 9)) %>%
  tibble::as_tibble()
names(temp_w) <- c("year", "N01", "N25", "N19") #data in "wide" format

head(temp_w)

(ggplot(data = temp_w,
        mapping = aes(x = year)) + 
    geom_line(mapping = aes(y = N01), color = "red") +
    geom_line(mapping = aes(y = N25), color = "blue") +
    geom_line(mapping = aes(y = N19), color = "green")
  ) #this method of plotting is not encouraged

temp_l <- tidyr::gather(temp_w, c("N01", "N25", "N19"), 
                        key = "Normal RV", 
                        value = "Normal Realization"
                        ) #converting data from wide to "long"

head(temp_l)

(ggplot(data = temp_l,
        mapping = aes(x = year, 
                      y = `Normal Realization`,
                      color = `Normal RV`)) +
    geom_line()
  ) #the preferred way of plotting is in the long format

```










## References